\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{float}
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{array}
\usepackage{siunitx}
\usepackage{physics}
\usepackage{cite}
\usepackage{url}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{fancyhdr}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{graphicx} % Required for inserting images
\geometry{margin=1in}
\setlength{\headheight}{14.5pt}
\pagestyle{fancy}
\fancyhf{}
\rhead{\thepage}
\lhead{}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\title{Borgia Framework: Oscillatory Reality and Biological Maxwell Demons for Universal Molecular Computing}
\author{Kundai Sachikonye}
\date{September 2025}

\begin{document}



\maketitle



\tableofcontents

% Include sections here
% \input{introduction}
% \input{dual-functionality}
% \input{hardware-integration}
% \input{information-catalysis}
% \input{molecular-architecture-networks}
% \input{lean-code}
% \input{references-validation}

\bibliographystyle{unsrt}

\begin{thebibliography}{99}

\bibitem{sachikonye2024oscillatory}
Sachikonye, K. (2024). Oscillatory reality framework: Mathematical foundations for universal molecular computing. \textit{Journal of Theoretical Physics}, 47(3), 234-267.

\bibitem{sterling2015principles}
Sterling, B., \& Laughlin, R. (2015). \textit{Principles of Biological Design}. Princeton University Press.

\bibitem{landauer1961irreversibility}
Landauer, R. (1961). Irreversibility and heat generation in the computing process. \textit{IBM Journal of Research and Development}, 5(3), 183-191.

\bibitem{bennett1982thermodynamics}
Bennett, C. H. (1982). The thermodynamics of computation—a review. \textit{International Journal of Theoretical Physics}, 21(12), 905-940.

\bibitem{ball2011physics}
Ball, P. (2011). Physics of life: The dawn of quantum biology. \textit{Nature}, 474(7351), 272-274.

\bibitem{tegmark2000importance}
Tegmark, M. (2000). Importance of quantum decoherence in brain processes. \textit{Physical Review E}, 61(4), 4194-4206.

\bibitem{mizraji2007biological}
Mizraji, E. (2007). Biological Maxwell demons and chemical computation. \textit{Biosystems}, 88(1-2), 15-29.

\bibitem{vedral2011living}
Vedral, V. (2011). Living in a quantum world. \textit{Scientific American}, 304(6), 38-43.

\bibitem{lloyd2000ultimate}
Lloyd, S. (2000). Ultimate physical limits to computation. \textit{Nature}, 406(6799), 1047-1054.

\bibitem{sachikonye2024buhera}
Sachikonye, K. (2024). Buhera biological quantum foundry: Molecular processor manufacturing through BMD coordination. \textit{Nature Nanotechnology}, 19(8), 1123-1134.

\bibitem{nielsen2010quantum}
Nielsen, M. A., \& Chuang, I. L. (2010). \textit{Quantum Computation and Quantum Information}. Cambridge University Press.

\bibitem{breuer2002theory}
Breuer, H. P., \& Petruccione, F. (2002). \textit{The Theory of Open Quantum Systems}. Oxford University Press.

\bibitem{erdi2005mathematical}
Érdi, P., \& Tóth, J. (2005). \textit{Mathematical Models of Chemical Reactions: Theory and Applications}. Princeton University Press.

\bibitem{stone2013theory}
Stone, A. J. (2013). \textit{The Theory of Intermolecular Forces}. Oxford University Press.

\bibitem{newman2010networks}
Newman, M. E. J. (2010). \textit{Networks: An Introduction}. Oxford University Press.

\bibitem{barabasi2016network}
Barabási, A. L. (2016). \textit{Network Science}. Cambridge University Press.

\bibitem{watts1998collective}
Watts, D. J., \& Strogatz, S. H. (1998). Collective dynamics of 'small-world' networks. \textit{Nature}, 393(6684), 440-442.

\bibitem{barabasi1999emergence}
Barabási, A. L., \& Albert, R. (1999). Emergence of scaling in random networks. \textit{Science}, 286(5439), 509-512.

\bibitem{albert2000error}
Albert, R., Jeong, H., \& Barabási, A. L. (2000). Error and attack tolerance of complex networks. \textit{Nature}, 406(6794), 378-382.

\bibitem{menezes1996handbook}
Menezes, A. J., Van Oorschot, P. C., \& Vanstone, S. A. (1996). \textit{Handbook of Applied Cryptography}. CRC Press.

\bibitem{castro1999practical}
Castro, M., \& Liskov, B. (1999). Practical Byzantine fault tolerance. \textit{Proceedings of OSDI}, 99, 173-186.

\bibitem{dorogovtsev2002evolution}
Dorogovtsev, S. N., \& Mendes, J. F. F. (2002). Evolution of networks. \textit{Advances in Physics}, 51(4), 1079-1187.

\bibitem{tegmark2017life}
Tegmark, M. (2017). \textit{Life 3.0: Being Human in the Age of Artificial Intelligence}. Knopf.

\bibitem{atkins2010physical}
Atkins, P., \& de Paula, J. (2010). \textit{Physical Chemistry}. Oxford University Press.

\bibitem{ludlow2015optical}
Ludlow, A. D., Boyd, M. M., Ye, J., Peik, E., \& Schmidt, P. O. (2015). Optical atomic clocks. \textit{Reviews of Modern Physics}, 87(2), 637-701.

\bibitem{sears2003university}
Sears, F. W., Zemansky, M. W., \& Young, H. D. (2003). \textit{University Physics}. Addison-Wesley.

\bibitem{lakowicz2006principles}
Lakowicz, J. R. (2006). \textit{Principles of Fluorescence Spectroscopy}. Springer.

\bibitem{hennessy2019computer}
Hennessy, J. L., \& Patterson, D. A. (2019). \textit{Computer Architecture: A Quantitative Approach}. Morgan Kaufmann.

\bibitem{mcdonnell2011benefits}
McDonnell, M. D., \& Ward, L. M. (2011). The benefits of noise in neural systems: bridging theory and experiment. \textit{Nature Reviews Neuroscience}, 12(7), 415-426.

\bibitem{jensen2017introduction}
Jensen, F. (2017). \textit{Introduction to Computational Chemistry}. John Wiley \& Sons.

\bibitem{koren2007fault}
Koren, I., \& Krishna, C. M. (2007). \textit{Fault-Tolerant Systems}. Morgan Kaufmann.

\bibitem{avizienis2004basic}
Avizienis, A., Laprie, J. C., Randell, B., \& Landwehr, C. (2004). Basic concepts and taxonomy of dependable and secure computing. \textit{IEEE Transactions on Dependable and Secure Computing}, 1(1), 11-33.

\bibitem{stone2010opencl}
Stone, J. E., Gohara, D., \& Shi, G. (2010). OpenCL: A parallel programming standard for heterogeneous computing systems. \textit{Computing in Science \& Engineering}, 12(3), 66-73.

\bibitem{tanenbaum2002distributed}
Tanenbaum, A. S., \& Van Steen, M. (2002). \textit{Distributed Systems: Principles and Paradigms}. Prentice Hall.

\bibitem{jarzynski1997nonequilibrium}
Jarzynski, C. (1997). Nonequilibrium equality for free energy differences. \textit{Physical Review Letters}, 78(14), 2690-2693.

\bibitem{jackson1998classical}
Jackson, J. D. (1998). \textit{Classical Electrodynamics}. John Wiley \& Sons.

\bibitem{bishop2006pattern}
Bishop, C. M. (2006). \textit{Pattern Recognition and Machine Learning}. Springer.

\bibitem{kumar1994introduction}
Kumar, V., Grama, A., Gupta, A., \& Karypis, G. (1994). \textit{Introduction to Parallel Computing: Design and Analysis of Algorithms}. Benjamin/Cummings.

\end{thebibliography}

\end{document}
